Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2023-12-29 mop/draft.tex
some edits to definition of targeting

\usepackage{enumitem}
\usepackage{tikz}
\setlist{nosep} % to reduce space in lists

@@

\usetikzlibrary{arrows.meta,positioning}
\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=1cm and 1cm, auto,
>=Latex,every node/.append style={align=center}, 
int/.style={draw, minimum size=1.6cm}] 
\node [int] (S) [right=of H] {$S$};
\node [int] (I) [right=of S] {$I$};
\node (Mcoord) [right=of I, coordinate] {Mcoord};
\node [int] (R1) [right=of Mcoord] {$R_1$};
\node [int] (R2) [right=of R1] {$R_2$};
\node [int] (R3) [right=of R2] {$R_3$};
\node [circle, draw] (M) [above=of Mcoord] {$M$};
\node (deathsS) [above=of S, coordinate] {deathsS};
\node (deathsI) [above=of I, coordinate] {deathsS};
\node (deathsR1) [above=of R1, coordinate] {deathsR1};
\node (deathsR2) [above=of R2, coordinate] {deathsR2};
\node (deathsR3) [above=of R3, coordinate] {deathsR3};
\node (belowR3) [below=of R3, coordinate] {belowR3};
\node (belowS) [below=of S, coordinate] {belowS};
\path[->]
    (H) edge node {$dH$} (S)
    (S) edge node {$\lambda(t)$}  (I)
    (I) edge node {$\gamma$}  (R1)
    (R1) edge node {$k\epsilon$}  (R2)
    (R2) edge node {$k\epsilon$}  (R3)
    (R3) edge node {} (belowR3) 
    (belowR3) edge node {$k\epsilon$} (belowS) 
    (belowS) edge node {} (S)
    (I) edge node {$m$}  (M)
    (S) edge node {$\delta$}  (deathsS)
    (I) edge node {$\delta$}  (deathsI)
    (R1) edge node {$\delta$}  (deathsR1)
    (R2) edge node {$\delta$}  (deathsR2)
    (R3) edge node {$\delta$}  (deathsR3)
\end{tikzpicture}
\caption{Illustration of SIR model from \cite{king08}.}
\label{fig:sir}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.53]{imgs/095/pairs.png}
    \includegraphics[scale=0.53]{imgs/095/qq.png}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. Left: Paired searches from the same starting point. Right: Q-Q plot of ranked IFAD searches against ranked IF2 searches. It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 5 log-likelihood units of it. We use a dotted red line to display the MLE in both. }
    \label{fig:scatter}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.52]{imgs/095/boxplot.png}
    \includegraphics[scale=0.54]{imgs/095/optim.png}
    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs. This simulates the result of the common procedure of running a few searches and choosing the best one. IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations. Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE. The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima. 
    In both, as before, we use a dotted red line to display the MLE. }
    \label{fig:boxplot-search}
\end{figure}

\begin{figure}[u]
    \centering
    \includegraphics[scale=0.7]{imgs/095/hist.png}
    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone. As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so. Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices. MOP alone performs poorly, demonstrating the necessity of IF2. }
    \label{fig:hist-all}
\end{figure}

@@

\begin{defn}[Targeting, \cite{liu1998smc}]
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any integrable function $h$,
    $$
    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
    $$

    A set of particles $(X_j, w_j), j=1,2, \ldots$, \textbf{targets} $\pi$ if
    $$
    \frac{\sum_{j=1}^J h(X_j) w_j}{\sum_{j=1}^J w_j} \stackrel{a.s.}{\to} E_\pi(h(X))
    $$
    for any integrable function $h$ as $J \to \infty$, in the sense of the triangular array convergence in \cite{chopin2004clt}. 
\end{defn}

It is known that the bootstrap and appropriately weighted particle filters target the posterior $f_{X_{1:N}|Y_{1:N}}$ in this sense, as long as they adhere to the procedure in \cite{chopin2004clt}. We now present three helper lemmas. The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem.

@@

\begin{lem}[Likelihood Proper Weighting]
    \label{lem:lik-proper-weight}
  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is properly estimated by either the before-resampling estimate,
\begin{equation}\label{L1}
L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
or by the after-resampling estimate,
\begin{equation}\label{L2}
L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
\end{equation}
where $L^\phi_n$ is as defined in the various algorithms.
\end{lem}

>>>>

\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

\setlist{nosep} % to reduce space in lists

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=1cm and 1cm, auto,
>=Latex,every node/.append style={align=center}, 
int/.style={draw, minimum size=1.6cm}]
  \node [int] (H) at (0,0) {$H$};
\node [int] (S) [right=of H] {$S$};
\node [int] (I) [right=of S] {$I$};
\node (Mcoord) [right=of I, coordinate] {Mcoord};
\node [int] (R1) [right=of Mcoord] {$R_1$};
\node [int] (R2) [right=of R1] {$R_2$};
\node [int] (R3) [right=of R2] {$R_3$};
\node [circle, draw] (M) [above=of Mcoord] {$M$};
\node (deathsS) [above=of S, coordinate] {deathsS};
\node (deathsI) [above=of I, coordinate] {deathsS};
\node (deathsR1) [above=of R1, coordinate] {deathsR1};
\node (deathsR2) [above=of R2, coordinate] {deathsR2};
\node (deathsR3) [above=of R3, coordinate] {deathsR3};
\node (belowR3) [below=of R3, coordinate] {belowR3};
\node (belowS) [below=of S, coordinate] {belowS};
\path[->]
    (H) edge node {$dH$} (S)
    (S) edge node {$\lambda(t)$}  (I)
    (I) edge node {$\gamma$}  (R1)
    (R1) edge node {$k\epsilon$}  (R2)
    (R2) edge node {$k\epsilon$}  (R3)
    (R3) edge node {} (belowR3) 
    (belowR3) edge node {$k\epsilon$} (belowS) 
    (belowS) edge node {} (S)
    (I) edge node {$m$}  (M)
    (S) edge node {$\delta$}  (deathsS)
    (I) edge node {$\delta$}  (deathsI)
    (R1) edge node {$\delta$}  (deathsR1)
    (R2) edge node {$\delta$}  (deathsR2)
    (R3) edge node {$\delta$}  (deathsR3);
\end{tikzpicture}
\caption{Illustration of SIR model from \cite{king08}.}
\label{fig:sir}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
    \includegraphics[scale=0.53]{../imgs/095/qq.png}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. Left: Paired searches from the same starting point. Right: Q-Q plot of ranked IFAD searches against ranked IF2 searches. It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 5 log-likelihood units of it. We use a dotted red line to display the MLE in both. }
    \label{fig:scatter}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.52]{../imgs/095/boxplot.png}
    \includegraphics[scale=0.54]{../imgs/095/optim.png}
    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs. This simulates the result of the common procedure of running a few searches and choosing the best one. IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations. Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE. The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima. 
    In both, as before, we use a dotted red line to display the MLE. }
    \label{fig:boxplot-search}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.7]{../imgs/095/hist.png}
    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone. As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so. Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices. MOP alone performs poorly, demonstrating the necessity of IF2. }
    \label{fig:hist-all}
\end{figure}

@@

\begin{defn}[Targeting]
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any integrable function $h$,
\begin{equation}
    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
\end{equation}  
    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if
\begin{equation}
    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
\end{equation}
for any integrable function $h$ as $J \to \infty$.
\end{defn}
\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS. ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS. I WONDER IF SOME OTHER REFERENCE, E.G. DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem.

@@

\begin{lem}[Likelihood Proper Weighting]
    \label{lem:lik-proper-weight}
  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
\begin{equation}\label{L1}
L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
or by the after-resampling estimate,
\begin{equation}\label{L2}
L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
\end{equation}
where $L^\phi_n$ is as defined in the various algorithms.
\end{lem}
